Data Management and Domain Adjusting Strategies for Implementing Parallel SPH Method in JPUE Simulation

### 2016-01-06 Paper for submission.tex
Wensday night

As for the actual storage of data representing the physical quantities associated to each particle, different strategies  have been adopted in existing implementations of SPH. 
In both SPHysics\cite{dominguez2011optimization} and DualSPHysics\cite{crespo2015dualsphysics}, The physical quantities of each particle (position, velocity, density…) are stored in arrays, and the particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is ease to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome.   Ferrari\cite{ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fix-size arrays are thereby also eliminated. We define C++ classes which contains all data of particle and bucket. As for the management of data, we adopt a hash table to store pointers to particles and buckets, which gives us not only flexibility of deleting and adding element, but also quicker access compared with linked list. Instead of using the "nature manner" to number particles, we adopt SFC based index to give each particle and background bucket an unique identifier -- a strategy known to preserve some locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time will have different identifiers.

@@

We will describe in this section   strategies which satisfies these requirements.

@@

Particles used in the test problem can be categorized into four types based the particle-type-flag: real particle, wall ghost particle, pressure ghost particle and eruption ghost particle. Ghost particles are for imposing of corresponding boundary conditions (see figure \ref{fig:Involved_initial}). As different types of particles involve different amount of computational work, shown by table \ref{tab:Computational_cost_particles} and table \ref{tab:Computational_cost_steps}, we assign different work load weight for different types of particles based on profilling data. Instead of simply using number of contained particles as work load for bucket, work load of each bucket is determined by summing up work load weight of all particles within the bucket. The SFC sequence passing through centroids of all buckets now becomes a weighted sequence. Domain decomposition  is conducted based on the weighted SFC of buckets.\\

@@

As simulation progresses, the ejected fluid will reach larger area and more and more particles will be influenced. When originally stationary air is influenced by erupted material, the mass fraction of the erupted material will increase from zero to a positive value. So we can determine whether a particle is involved or not based on whether the mass fraction of that particle is larger than a given threshold ($10^{-5} $ in our simulation). Other physical properties, such velocity, can also serve as alternative ``swich criteria".\\

@@

\caption{ Adjusting the domain based on Involved flag of particles}

@@

A similar scheme can be deployed for buckets too resulting in a categorization  of buckets. The additional complexity here is that potentially involved ghost particles must also be accounted for.

@@

\begin{eqnarray}

@@

\includegraphics[scale=0.33]{weak_loglog}

@@

Experiments have been carried out on the computational cluster of Center for Computational Research (CCR) at Buffalo. Two types of processors -- 8 core XXXXX running at YYY
clock speed with ZZZ memory per core and 12 core AAAAA running at BBBB
clock speed with CCC memory per core. Each node comprised of two sockets with two of these processors. Memory and level 3 cache are shared on each node.

@@

The initial domain is $[-4800m,4800m]\times [-4800m,4800m] \times [0m, 6000m]$, with smoothing length (we set initial intervals between particles equal to smoothing length) equals to 200m and 100m respectively for test case1 and test case2, respectively. The computational work load of test case 2 is 8 times of that of the test case 1. The simulations run for 20s physical time.  Wall clock time and speed up are shown in figure \ref{fig:2cases_time} and \ref{fig:2cases_efficiency}. Linear speed up is observed when number of processes is smaller than 16. Test case2 shows better speed up than test case1 which implies that the overhead of strong scalability can be increased by increasing total amount of work load. We also compared the performance of 2D domain decomposition and 3D domain decomposition. As shown by figure \ref{fig:2D_vs_3D_efficiency}, 2D domain decomposition shows a little bit better speed up than 3D domain decomposition for 32 processes. This can be explained by the fact that 2D domain decomposition can get more regular subdomain and as a results, need less communication. When number of processes increase, it becomes harder for 2D decomposition to balance workload, that is why 3D domain decomposition has better speed up for processes number equals to 64 and 128. The weak scalability test is conducted with the same initial domain and various smoothing length. Each simulation runs for 400 loops. The average number of real particles of each process keep constant at $259000$. As shown in figure \ref{fig:weak_loglog}, simulation time increases pretty slowly when number of processes is larger than 132. the relative sharp increase in simulation time from 32 processes to 132 processes might be caused by an increase of number of neighbor processes. 
%That can be explained by the fact that the total number of footprint buckets in x-y plane is 64 in the test case.
%When we request 128 processes, 64 processes will be totally idle at the beginning. As simulation goes, the domain will expand and more and more idle processes will start getting tasks. That's why we still get some speed up when number of process is 128. 3D domain decomposition is able to divide the work load in a more flexible way and achieve better strong scalability.

@@

\section{Conclusion}
We developed data management strategies for parallel implementation of SPH method using MPI standard to simulate complicated problems, such as JUPE, which requires flexible and fast data retrievalling, adding and deleting. Neighbors searching and domain decomposition is based on background grid which overlaps the domain and keep stationary during simulation. SFC based index scheme, which provides a global numberring methodology which is purely coordinates dependent, is adopted to give each bucket an unique identifier. A time dependent key which is also based on SFC is used as identifier for particle. 
Hashtables with external link list are adopted for accessing particles and buckets data. Based on weighted particle work load, a dynamic load balance strategy is developed by checking load balance and redecomposing the domain at an optimize interval. The performance of the code was further improved to several times faster by adjusting computational domain according to progress of simulation. 
Scalability tests on our code shew acceptable strong scalability and good weak scalability. 3D domain decomposition provided better strong scalability than 2D domain decomposition when number of processes is larger.
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captions off option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
%
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure}

>>>>

As for the actual storage of data representing the physical quantities associated to each particle, different strategies  have been adopted in existing implementations of SPH. 
In both SPHysics\cite{dominguez2011optimization} and DualSPHysics\cite{crespo2015dualsphysics}, The physical quantities of each particle (position, velocity, density…) are stored in arrays, and the particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is ease to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome.   Ferrari\cite{ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fix-size arrays are thereby also eliminated. We define C++ classes which contains all data of particle and bucket. As for the management of data, we adopt a hash table to store pointers to particles and buckets, which gives us not only flexibility of deleting and adding element, but also quicker access compared with linked list. Instead of using the "nature manner" to number particles, we adopt SFC based index to give each particle and background bucket an unique identifier -- a strategy known to preserve some locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time will have different identifiers. 

@@

We will describe in this section strategies which satisfy these requirements.

@@

Particles used in the test problem can be categorized into four types based on the particle-type-flag: real particle, wall ghost particle, pressure ghost particle and eruption ghost particle. Ghost particles are for imposing of corresponding boundary conditions (see figure \ref{fig:Domain_adjusting}). As different types of particles involve different amount of computational work, shown by table \ref{tab:Computational_cost_particles} and table \ref{tab:Computational_cost_steps}, we assign different work load weight for different types of particles based on profilling data. Instead of simply using number of contained particles as work load for bucket, work load of each bucket is determined by summing up work load weight of all particles within the bucket. The SFC sequence passing through centroids of all buckets now becomes a weighted sequence. Domain decomposition  is conducted based on the weighted SFC of buckets.\\

@@

As simulation progresses, the ejected fluid will reach larger area and more and more particles will be influenced. When originally stationary air is influenced by erupted material, the mass fraction of the erupted material will increase from zero to a positive value. So we can determine whether a particle is involved or not based on whether the mass fraction of that particle is larger than a given threshold ($10^{-5} $ in our simulation). Other physical properties, like velocity, can also serve as alternative "swich criteria".\\

@@

\caption{ Adjusting the domain based on involved-flag of particles}

@@

A similar scheme can also be deployed for buckets resulting in a categorization  of buckets. The additional complexity here is that potentially involved ghost particles must also be accounted for.

@@

\begin{eqnarray} %%I did not notice that I can add equation like this, useful, haha

@@

\includegraphics[scale=0.35]{weak_loglog}

@@

Experiments have been carried out on the computational cluster of Center for Computational Research (CCR) at Buffalo. Two types of processors -- 8 core L5630 running at 2.13GHz clock rate with 3GB memory per core on a Mellanox network and 12 core E5645 running at 2.40GHz clock rate with 4GB memory per core on a Q-Logic network. Each node comprised of two sockets with two of these processors. Memory and level 3 cache are shared on each node.

@@

The initial domain is $[-4800m,4800m]\times [-4800m,4800m] \times [0m, 6000m]$, with smoothing length (we set initial intervals between particles equal to smoothing length) equals to 200m and 100m respectively for test case1 and test case2, respectively. The computational work load of test case 2 is 8 times of that of the test case 1. The simulations run for 20s physical time.  Wall clock time and speed up are shown in figure \ref{fig:2cases_time} and \ref{fig:2cases_efficiency}. Linear speed up is observed when number of processes is smaller than 16. Test case2 shows better speed up than test case1 which implies that the overhead of strong scalability can be increased by increasing total amount of work load. We also compared the performance of 2D domain decomposition and 3D domain decomposition. As shown by figure \ref{fig:2D_vs_3D_efficiency}, 2D domain decomposition shows a little bit better speed up than 3D domain decomposition for 32 processes. This can be explained by the fact that 2D domain decomposition can get more regular subdomain and as a results, need less communication. When number of processes increase, it becomes harder for 2D decomposition to balance workload, that is why 3D domain decomposition has better speed up for processes number equals to 64 and 128. The weak scalability test is conducted with the same initial domain and various smoothing length. Each simulation runs for 400 loops. The average number of real particles of each process keep constant at $25900$. As shown in figure \ref{fig:weak_loglog}, simulation time are almost constant with some minor fluctutations around the average value. These fluctuations are due to sharing nodes with other users. In addition, difference in simulation time for different processors are obvious.
%That can be explained by the fact that the total number of footprint buckets in x-y plane is 64 in the test case.
%When we request 128 processes, 64 processes will be totally idle at the beginning. As simulation goes, the domain will expand and more and more idle processes will start getting tasks. That's why we still get some speed up when number of process is 128. 3D domain decomposition is able to divide the work load in a more flexible way and achieve better strong scalability.

@@

\section{Conclusion}
We developed data management strategies for parallel implementation of SPH method using MPI standard to simulate complicated problems, such as JUPE, which requires flexible and fast data retrievalling, adding and deleting. Neighbors searching and domain decomposition is based on background grid which overlaps the domain and keep stationary during simulation. SFC based index scheme, which provides a global numberring methodology which is purely coordinates dependent, is adopted to give each bucket an unique identifier. A time dependent key which is also based on SFC is used as identifier for particle. 
Hashtables with external link list are adopted for accessing particles and buckets data. Based on weighted particle work load, a dynamic load balance strategy is developed by checking load balance and redecomposing the domain at an optimize interval. The performance of the code was further improved to several times faster by adjusting computational domain according to progress of simulation. 
Scalability tests on our code shew acceptable strong scalability and good weak scalability. 3D domain decomposition provided better strong scalability than 2D domain decomposition when number of processes is larger.
% An example of a floating figure using the graphicx package.
% Note that \label must occur AFTER (or within) \caption.
% For figures, \caption should occur after the \includegraphics.
% Note that IEEEtran v1.7 and later has special internal code that
% is designed to preserve the operation of \label within \caption
% even when the captions off option is in effect. However, because
% of issues like this, it may be the safest practice to put all your
% \label just after \caption rather than within \caption{}.
% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
% option should be used if it is desired that the figures are to be
% displayed while in draft mode.
%\begin{figure}[!t]
%\centering
%\includegraphics[width=2.5in]{myfigure}
% where an .eps filename suffix will be assumed under latex, 
% and a .pdf suffix will be assumed for pdflatex; or what has been declared
% via \DeclareGraphicsExtensions.
%\caption{Simulation results for the network.}
%\label{fig_sim}
%\end{figure}
